On the reality of the quantum state once again

### 2021-12-01 Articles/2021-HarriganSpekkens/HarriganSpekkens.tex
HarriganSpekkens: another pass and added discussion

Let us now step back a moment. While the math is clear enough, we still need to understand what is going on at a conceptual level. Why does assuming $p(\lambda|P)$ lead to an epistemic structure incompatible with the quantum one? To answer this question, let us review how probability spaces work in both classical and quantum mechanics.

@@

This procedure is repeated for all observables. The probability measure of each sums to one over the respective sub-algebra, over the respective basis. There are infinitely many observables (e.g. each direction of spin) whose bases are disjoint sets. If we added all measures over the entire $\sigma$-algebra (e.g. all probabilities over all directions of spin) the sum would be infinite. In other words: we do not have a single probability distribution on a single $\sigma$-algebra. For each observable, we a probability distribution defined on a particular sub-algebra\footnote{This is exactly why some rules of classical probability are not allowed in quantum probability: we can only combine statements within the same sub-algebra and not across different sub-algebras.} and this is what gives rise to contextuality. Before getting a probability distribution we need to specify a context, a sub-algebra, a basis.

We may be tempted to think that we have one context before and one after the measurement, but this does not quite work. Each context is needed, for example, to define the expectation value of the related observables. It is by measuring those quantities over an ensemble of identically prepared system (i.e. quantum tomography) that we can reconstruct the initial state. Therefore we need to be able to say that all distributions over all contexts are fixed by the preparation. The measurement process can only measure one without perturbing the other.

Yet, at each time there is a special context. If we take the $z+$, the probability distribution along that direction will the sharpest while the probability distribution along a perpendicular direction will be uniform. Therefore there is a context in which the entropy is minimal, there is a context that gives us the most information about the system. This is the basis over which we calculate the entropy of a mixed state, this is the basis in which the density matrix diagonalizes.

Note that the minimal entropy context changes over mixes. For an equal mix of $z+$ and $x+$, the minimal entropy distribution is neither over $Z$ or $X$. This is why the epistemic structure of contextual and non-contextual models is so different: classical mixtures affect the context.

To sum up, assuming $p(\lambda|P)$ exists means assuming there is a single probability space, a single $\sigma$-algebra, a single set of propositions, to which all probabilities are assigned. This is exactly assuming non-contextuality. To allow for contextuality we should assume that for each observable $O$ we have a probability $p_O(\lambda_O|P)$ of having prepared one of the eigenstates, such that $\sum_{\lambda_O} p_O(\lambda_O|P) = 1$. However, the sum over all states $\sum_{\lambda} p_{O(\lambda)}(\lambda|P) = + \infty$ diverges and is not a probability measure. If it didn't diverge, then we could renormalize and put a probability measure over the whole space, making the model non-contextual. Mathematically, the probability space is not Kolmogorov: probability of countable sequences does not always converge.\footnote{In our larger project, we are trying to understand what are the correct axioms for such space with mixed success.}

@@

>>>>

Let us now step back a moment. Why does assuming $p(\lambda|P)$ lead to an epistemic structure incompatible with the quantum one?

It is widely understood that one feature of quantum mechanics is non-contextuality: a measurement cannot be understood as revealing pre-existing values. We may be tempted to think that we have one context before and one after the measurement. What we are seeing that this does not work because we a dual effect on preparations. As we cannot understand quantum measurements as selecting from a single joint probability distribution, we cannot understand quantum preparations as constructing a single joint probability distribution. To fully understand how this works, let us review how how probability spaces work in both classical and quantum mechanics.

@@

Mathematically, what we call context is a $\sigma$-algebra: a set of propositions upon which we assign probabilities. This tells us why we cannot simply use classical rules in quantum mechanics. Classical mechanics has only one context, so we can always mix and match statements however we please. In quantum mechanics, propositions about different observables live in different $\sigma$-algebras, in different probability spaces, and therefore cannot be combined.

One may think that we can get away by assuming that, at each time, there is a ``true'' context. And this seems to work in some cases. Given any density matrix $\rho$, there exist a (non-unique) privileged context: this is the one in which we can express $\rho = \sum p_i |\psi_i > < \psi_i|$ as the mix of orthogonal pure states. If $\rho$ is itself a pure state, any basis that includes it will be a privileged context. This privileged context provides us the probability distribution over which the information entropy is calculated. After a measurement, the privileged context changes to one that includes the eigenstates of the chosen observable. However, this does not work in general.

Suppose we start with two density matrixes $\rho_1$ and $\rho_2$ with different eigenstates. We then create the mixture $\rho = p_1 \rho_1 + p_2 \rho_2$. The contexts of the mixture will not be the same as the original contexts. It will, in fact, even depend on $p_1$, $p_2$. That is, when mixing preparations we are not simply combining two probability distribution over the same context, like in classical mechanics, but we are combining the contexts themselves. Non-contextuality does not affect only measurement, but, dually, preparations as well. And this affect how information entropy is calculated and the overall epistemic structure.

% TODO: mention tomography?

To sum up, assuming $p(\lambda|P)$ exists means assuming there is a single probability space, a single $\sigma$-algebra, a single set of propositions, to which all probabilities are assigned. This is exactly assuming non-contextuality. While this can be made to work for pure state to pure state transitions during measurement, mixing preparations leads to an epistemic structure that is fundamentally non-contextual, leading to inconsistencies with quantum mechanics. To allow for contextuality we should assume that for each observable $O$ we have a probability $p_O(\lambda_O|P)$ of having prepared one of the eigenstates, such that $\sum_{\lambda_O} p_O(\lambda_O|P) = 1$. The entropy of the preparation would be the lowest entropy across all contexts.

@@

\section*{Supplementary material}

TODO: do we need to bump anything here?
